---
title: "Fitting a spatiotemporal model with a GAM and deriving an abundance index"
date: "`r format(Sys.Date(), format='%B %d %Y') `"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a spatiotemporal model and deriving an abundance index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, message=FALSE, error=FALSE, warning = FALSE, 
  comment = "#>"
)
```

> Adapted from Lewis Barnett's sdmTMB tutorial for the Fisheries Survey Course at UW: [GitHub pbs-assess/sdmTMB-teaching 2023-02-14](https://github.com/pbs-assess/sdmTMB-teaching/blob/490ee04157f19d1059c55a1318b894e8b8f460bb/uw-survey-2023/02-exercise.qmd#L356)

# Goals of the original exercise

-   Practice fitting a basic spatiotemporal model.
-   Understand how to inspect the model output.
-   Practice predicting from the model on new data and making visualizations of those predictions.
-   Gain familiarity with fitting, comparing and interpreting different random field structures.
-   Calculate an area-weighted biomass index and compare how model structure can impact an index.

# Our goals for this vignette:

-   Fit a GAM and obtain an area-weighted biomass index
-   Fit a basic spatiotemporal model (a GLMM!) using sdmTMB and compare GAM to GLMM

```{r install_inla, eval = FALSE, include = FALSE}
# install.packages("INLA", repos = c(getOption("repos"), INLA = "https://inla.r-inla-download.org/R/stable"), dependencies = TRUE)
#library(remotes)
#remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
```

```{r}
PKG <- c(
  "sdmTMB", # install.packages("sdmTMB", dependencies = TRUE)
  "mgcv", 
  "gratia",
  "visreg", 
  "gstat",
  "dplyr", 
  "ggplot2", 
  "INLA",
  "prediction",
  "inlabru", 
  "purrr")

for (p in PKG) {
  if(!require(p,character.only = TRUE)) {  
    install.packages(p)
    require(p,character.only = TRUE)}
}

options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
theme_set(theme_light())
```

Check your system:

```{r}
sessionInfo()
```

# The data

We will work with CPUE data from yellowfin sole in the Eastern Bering Sea summer bottom trawl survey. This data is publicly available through FOSS.

```{r, include = FALSE}
# Load EBS yellowfin sole CPUE table for example:
# load(here::here("data/ebs_yfs_example.rda"))
# dat <- ebs_yfs_example
# If the ebs_yfs_example.rds file isn't available:
# load(here::here("data/noaa_afsc_public_foss.rda"))
dat <- yfs <- sdmgamindex::noaa_afsc_public_foss |> 
  dplyr::filter(srvy=="EBS" & species_code == 10210)
# saveRDS(object = yfs,file = here::here("data/ebs_yfs_example.RDS"))

```

## Add UTM columns, log depth, and year as factor

```{r}
dat <- add_utm_columns(dat, 
                       ll_crs = 4326,
                       ll_names = c("longitude_dd_start", "latitude_dd_start"))

dat$log_depth <- log(dat$depth_m)
dat$year_factor <- as.factor(dat$year)

dat[,c("X","Y")]
range(dat$X)
range(dat$Y)

ggplot(dat, aes(X, Y, size = cpue_kgkm2)) +
  geom_point(shape = 21) +
  coord_fixed()
```

The dataset contains sampling locations (`longitude` and `latitude`) and `year`. It also contains sampling `depth` in meters and sample CPUE `density` in units of tonnes/km^2^.

-   <div>

    ```{r}
    ggplot(data = dat, 
           mapping = aes(x = longitude_dd_start, y = latitude_dd_start, 
                         size = cpue_kgkm2, color = bottom_temperature_c)) + 
      geom_point(alpha = 0.3)
    ```

# The prediction grid
## Load the grid

Extrapolation grid for the EBS. Ignore the warnings about the UTM zones-- it's a Bering Sea thing.

```{r}
# This is another form of the same grid:
load(here::here("data/pred_grid_ebs.rda")) # object: pred_grid_ebs
#pred_grid_ebs <- read.csv(here::here("data/ebs_2022_epsg3338.csv"),header = TRUE)

get_crs(dat = pred_grid_ebs,ll_names =c("lon","lat"))

grid <- add_utm_columns(pred_grid_ebs, 
                       #ll_crs = 32603, 
                       ll_names = c("lon", "lat"))
grid$area_swept_km2 <- 1 # if you have area swept in your model as an offset (common when you're using CPUE data) you can do this (I think)

range(grid$X)

grid$Shape_Area_ha <- grid$Shape_Area*0.0001 # original Shape_area is in m^2

```
## Replicate the grid for each year

When you have spatiotemporal data, you need a grid for each year. There is a nice tidy little chunk of `{purrr}` code that will do that for you!

```{r}
grid <- purrr::map_dfr(unique(dat$year), ~ tibble(grid, year = .x))
```


# Fit a GAM to spatial data (analogous to spatial-only model in sdmTMB)
## Temporal effect plus spatial smoother, no covariates

First, fit a GAM without any covariates: just year as a factor and a spatial smoother $s(X,Y)$. The `as.factor(year)` part is a common component of SDMs that are being used to generate indices.

```{r}
start.time <- Sys.time()
fit_gam <- gam(
  formula = cpue_kgkm2 ~ as.factor(year) +
    s(X, Y, k = 50),
  family = tw(link = "log"),
  data = dat
)

cat(
  "The GAM took ",
  difftime(Sys.time(), start.time, units = "mins"),
  " mins to run"
)
```

Get diagnostics and perform model checking:

```{r}
gam.check(fit_gam)
```

Include a 2-D smooth over space, and depth as a covariate.

```{r}
start.time <- Sys.time()
fit_gam_s <- gam(
  formula = cpue_kgkm2 ~ s(depth_m) + as.factor(year) + 
    s(X,Y), 
  family = tw(link = "log"),
  data = dat
)

cat("The GAM took ", 
    difftime(Sys.time(),start.time, units = "mins"), 
    "mins to run")

gam.check(fit_gam_s)
```

## Spatiotemporal effect

**NOTE**: This takes a long time to fit. Include a 2-D smooth over space for each year. In this case, there a different 2-D smooth for each year. This will take much longer to fit than the other two (this model takes about 10-15 mins - I have set `cache=TRUE` in the chunk so that the outputs are saved and can be used again the next time this doc is run). This model also includes area swept as an offset.

```{r cache=TRUE}
#| cache = TRUE
start.time <- Sys.time()

fit_gam_st <- gam(
  formula = count ~ as.factor(year) +  s(X, Y) +
    s(X, Y, by = as.factor(year)) + offset(area_swept_km2),
  family = tw(link = "log"),
  data = dat
)

cat(
  "The GAM took ",
  difftime(Sys.time(), start.time, units = "mins"),
  "mins to run"
)

```

Get diagnostics and perform model checking:

```{r}
gam.check(fit_gam_st)

# To save the k' values, edf, k-index, and p-value:
qq.gam(fit_gam_st, rep = 0, level = 0.9, rl.col = 2, 
       rep.col = "gray80")
```

Review console output to help verify convergence, and whether there were an adequate number of basis functions (k). If you compare this to the `gam.check()` results from the GAM above with no spatiotemporal effects, you can see that the Q-Q plot is closer to the line.

Examine the four diagnostic plots. Each of these gives a different way of looking at your model residuals. On the top-left is a Q-Q plot, which compares the model residuals to the expected/assumed distribution family. A well-fit model's residuals will be close to the 1-1 line, otherwise there may be under- or over-dispersion present. On bottom left is a histogram of residuals. We want this to have a shape similar to the distribution family we specified. On top-right is a plot of residual values as a function of the linear predictor. These should be evenly distributed around zero in a well-fitted model. Finally, on the bottom-right is plot of response against fitted values. A well-fitted model would show values near the 1-1 line.

# Calculate biomass index from GAM via simulation
## Predict to full survey area with new data
### Predict CPUE across the grid

To get predicted CPUE across the survey grid. Here you can use any of the models from above, as long as they have `YEAR` as a factor and as long as all your covariates are also in the grid. Here we use the basic model (`fit_gam`) because `grid` does not contain depth.

```{r}
pred_gam <- predict(fit_gam, type = "response", newdata = grid) #This takes a long time
pred_gam_df <- cbind(grid, pred_gam)
pred_gam_df$predicted_tot_grid <- pred_gam_df$Shape_Area_ha*pred_gam_df$pred_gam

```

Plot predictions over survey area. Note: Because the EBS grid is irregularly spaced, you have to use `geom_point()` here instead of `geom_tile()` or `geom_raster()`.

```{r}
pred_gam_df |>
  dplyr::filter(year == 2019) |> # single year
  ggplot(aes(X, Y, color = pred_gam)) +
  geom_point(size = 0.1) +
  scale_fill_viridis_c() +
  # facet_wrap(~year) +
  coord_fixed() +
  labs(color = expression(CPUE(kg / km^2))) +
  theme_light() +
  theme(legend.position = "bottom")
```

## Calculate index

Generally, model-based indices of abundance use the area of each grid cell and the prediction for that area to calculate a total index across the survey area. The simplest way to do this is to sum the predicted biomasses across the full grid like so:

```{r}
gam_idx_mt <- pred_gam_df |> 
  dplyr::group_by(year) |> 
  summarize(total_wt_mt = sum(predicted_tot_grid)/1000)

head(gam_idx_mt)
```

## Get uncertainties for GAM-based index by sampling from posteriors

[`{gratia}`](https://gavinsimpson.github.io/gratia/) is an R package for evaluating and displaying GAM fits. This script uses the `fitted_samples()` function, which draws fitted values from the posterior of a model using a Gaussian approximation. This will also take a while.

```{r}
sims <- gratia::fitted_samples(fit_gam, n=10, data=grid, 
                               scale="response", seed=9)
sims$year <- grid$year[sims$row]
sims$area <- rep(pred_gam_df$Shape_Area_ha, times = 10) # matching the # of draws
sims$biomass <- sims$fitted * sims$area # expand from density to biomass, given area

level <- 0.95 # specify probability for confidence interval

# Get sum of simulated biomass (density*area) across grid cells, with CI
lwr_fn <- function(x) {as.numeric(quantile(x, probs = (1 - level) / 2))}
upr_fn <- function(x) {as.numeric(quantile(x, probs = 1 - (1 - level) / 2))}

sims_sum <- sims %>% 
  group_by(year, draw) %>% 
  summarise_at("biomass", list(biomass = sum)) %>%
  group_by(year) %>%
  summarise_at("biomass", list(est = median, # could use mean
                           lwr = lwr_fn,
                           upr = upr_fn))
```

## Plot the index

Now you can plot your GAM model-based index over time just like we do for all the other indices!

```{r}
p1 <- sims_sum |>
  ggplot() +
  geom_line(aes(x = year, y = est)) +
  geom_ribbon(aes(x = year, y = est, ymin = lwr, ymax = upr), alpha = 0.4) +
  xlab("Year") +
  ylab("Biomass estimate (tons)")

p1
```

# Fit an analogous model in `` `sdmTMB` ``

The spatiotemporal GAM we fit above (`fit_gam_st`) should have an analog in sdmTMB. Here I am trying to get the same (more or less) index using sdmTMB instead.

To calculate an index from a model in `sdmTMB`, we need to run the `predict.sdmTMB()` function with the argument `return_tmb_object = TRUE`. We can then run the `get_index()` function to extract the total biomass calculations and standard errors.

We can set the area argument to our `cell_area` column in km^2^. In this case the value is 4 km^2^ for all of the cells, since our grid cells are 2 km x 2 km. If some grid cells were not fully in the survey domain (or were on land), we could feed a vector of grid areas to the area argument that matched the number of grid cells. Because the density units are tonnes per km^2^ for this data, the index is in tonnes.

```{r cache = TRUE}

# make a mesh for the predictions - this takes a while with cutoff = 10!!
mesh <- make_mesh(dat, c("X", "Y"), cutoff=10)
plot(mesh)

fit <- sdmTMB(data = dat, 
              formula = cpue_kgkm2 ~ 1 + as.factor(year), # If you set the intercept to zero, R spontaneously aborts!
              mesh = mesh,
              family = tweedie(link = "log"),
              time= "year"
              )

sanity(fit)
tidy(fit)

p <- predict(fit, newdata = grid, return_tmb_object = TRUE)

grid$cell_area <- 4 #Shape_Area_ha
index <- get_index(p, area = grid$Shape_Area_ha, bias_correct = FALSE)

ggplot(index, aes(year, est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  xlab("Year") +
  ylab("Biomass estimate (tonnes)")
```

We used `bias_correction = FALSE` to speed things up, but for any final result you will want to use the bias correction. Let's see how much the scale of the index changes with bias correction.

```{r}
index_c <- get_index(p, area = grid$cell_area, bias_correct = TRUE)
index_c$Method <- "Bias correction"

bind_rows(index, index_c) %>%
  ggplot(aes(year, est, fill = Method)) +
  geom_line(aes(colour = Method)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  xlab("Year") +
  ylab("Biomass estimate (tonnes)")
```

# 
