---
title: "Comparing tinyVAST to GAM-based indices: An example with GOA POP"
date: "`r format(Sys.Date(), format='%B %d %Y') `"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tinyVAST vs GAMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, message = FALSE, error = FALSE, warning = FALSE,
  comment = "#>"
)

# Index and mapping
#library(sdmgamindex)
library(dplyr)
#library(sf)
library(gstat)
#library(raster)
library(akgfmaps) # devtools::install_github("afsc-gap-products/akgfmaps", build_vignettes = TRUE)
library(coldpool) # devtools::install_github("afsc-gap-products/coldpool")
library(ggplot2)

# VAST tools
library(tinyVAST) # devtools::install_github("vast-lib/tinyVAST", dependencies = TRUE)
library(fmesher) # for making pred meshes for tinyVAST

# Aesthetics
options(ggplot2.continuous.colour = "viridis")
options(ggplot2.continuous.fill = "viridis")
theme_set(theme_bw())
```

# Pollock in the Gulf of Alaska

For this vignette, we will fit a tinyVAST model and a GAM to data for GOA Pacific ocean perch and compare the total biomass indices. 

### Load public catch data
```{r}
SPECIES <- c("Pacific ocean perch")
YEARS <- seq(2015,2023, by =2) # odd years for GOA survey
nyrs <- length(YEARS)
SRVY <- "GOA"

# Public catch data
dat0 <- sdmgamindex::noaa_afsc_public_foss %>% 
  dplyr::filter(srvy == "GOA" &
                  year %in% YEARS &
                  common_name %in% SPECIES) %>%
  dplyr::mutate(hauljoin = paste0(stratum, "_", station, "_", date_time)) %>%
  dplyr::select(
    year, date_time, latitude_dd_start, longitude_dd_start,
    cpue_kgkm2, common_name, 
    bottom_temperature_c, depth_m, # covariates (used later)
    srvy, area_swept_km2, duration_hr, vessel_id, hauljoin
  )

head(dat0)
```



## GAM

For tinyVAST, the data need to be in a specific format. Let's format it properly:

```{r}
dat <- dat0 |>
  dplyr::rename('x' = longitude_dd_start,
                'y' = latitude_dd_start,
                'time' = year) |>
                dplyr::mutate(catch_kg = cpue_kgkm2 * area_swept_km2
                  )
head(dat)
#save(dat, file = "FOSS_POP.Rdata")

```

### Build prediction grid/mesh

```{r}
mesh <- fm_mesh_2d(dat[,c('x','y')])
plot(mesh)
```

### Fit model
```{r}
vastfit <- tinyVAST(formula = cpue_kgkm2 ~ 0 + factor(time), #+ offset(area_swept_km2),
                    data = dat, dsem = NULL,
                    spatial_graph = mesh,
                    family = tweedie()
                    )
vastfit
```

### Get area-weighted total abundance
Use `tinyVAST()` to get an area-weighted total abundance. First we need to get an extrapolation grid; we'll use the one that the ModSquad group uses for model-based indices for GOA.

#### Prediction grid
```{r}
extrap_grid <- read.csv("../data/pred_grid_goa.csv")
plot(extrap_grid$Longitude,extrap_grid$Latitude)
extrap_grid_years <- sdmTMB::replicate_df(dat = extrap_grid,time_name = "time",time_values = unique(dat$time))
```



```{r}
# Test  the integrate_output() function
integrate_output(object = vastfit, 
                 newdata = subset(extrap_grid_years, time == 2019), 
                 bias.correct = FALSE) # bias.correct=FALSE means no random effects

# Apply it across all the years to get an index value
Est <- sapply(YEARS, FUN = \(t) integrate_output(
  object = vastfit,
  newdata = subset(extrap_grid_years, time == t),
  bias.correct = FALSE
))
colnames(Est) <- YEARS

# Design-based
# Pull db estimates from GAP_PRODUCTS
library(RODBC)
channel <- gapindex::get_connected()
dbdat <- RODBC::sqlQuery(channel = channel,
                       query = "
-- Select columns for output data
SELECT
SURVEY_DEFINITION_ID,
BIOMASS_MT / 1000000 AS BIOMASS_MMT,
(BIOMASS_MT - 2 * SQRT(BIOMASS_VAR)) / 1000000 AS BIOMASS_CI_DW,
(BIOMASS_MT + 2 * SQRT(BIOMASS_VAR)) / 1000000 AS BIOMASS_CI_UP,
BIOMASS_VAR,
YEAR

-- Identify what tables to pull data from
FROM GAP_PRODUCTS.AKFIN_BIOMASS

-- Filter data results
WHERE SPECIES_CODE = 30060
AND SURVEY_DEFINITION_ID = 47
AND AREA_ID = 99903
AND YEAR BETWEEN 1990 AND 2023" ) 
dbdat <- subset(dbdat, YEAR %in% colnames(Est))

# tinyVAST
#dev.off()
plot(YEARS, Est[1, ], type = "b", ylab = "Estimated biomass", pch=19, ylim = c(1e8,2e8))
arrows(x0 = YEARS, x1 = YEARS, y0 = Est[1,]-Est[2,],
       y1 = Est[1,]+Est[2,],angle=90,code = 3, length=0.05)

# Add design-based
points(dbdat$YEAR+0.1, dbdat$BIOMASS_MMT*1e8, pch=19,col = "blue")
arrows(x0 = YEARS+0.1, x1 = YEARS+0.1, y0 = dbdat$BIOMASS_CI_DW*1e8,
       y1 = dbdat$BIOMASS_CI_UP*1e8,
       angle=90,code = 3, length=0.05, col = 'blue')

```

## Fit GAM using `sdmgamindex`

### Set up the data for sdmgamindex:
```{r}
# project spatial data
crs_proj <- "EPSG:3338" # NAD83 / Alaska Albers
crs_latlon <- "+proj=longlat +datum=WGS84" # decimal degrees

ll <- sdmgamindex::convert_crs( 
  x = dat$longitude_dd_start,
  y = dat$latitude_dd_start, 
  crs_in = crs_latlon, 
  crs_out = crs_proj) 

YEARS <- sort(unique(dat$year))

# The sdmgamindex::get_surveyidx() expects some columns to be named in a specific way
dat_wrangled <- dat %>% 
  dplyr::rename(
    Year = year,
    wCPUE = cpue_kgkm2, 
    COMMON_NAME = common_name,
    GEAR_TEMPERATURE = bottom_temperature_c, 
    BOTTOM_DEPTH = depth_m,
    HaulDur = duration_hr,
    EFFORT = area_swept_km2,
    Ship = vessel_id) %>%
  dplyr::mutate( 
    # create some other vars
    Lon = longitude_dd_start, 
    Lat = latitude_dd_start, 
    lon = ll$X,
    lat = ll$Y,
    sx = ((longitude_dd_start - mean(longitude_dd_start, na.rm = TRUE))/1000),
    sy = ((latitude_dd_start - mean(latitude_dd_start, na.rm = TRUE))/1000), 
    ctime = as.numeric(as.character(Year)),
    date_time = as.Date(x = date_time, format = "%m/%d/%Y %H:%M:%S"), 
    hour = as.numeric(format(date_time,"%H")),
    minute = as.numeric(format(date_time,"%M")),
    day = as.numeric(format(date_time,"%d")),
    month = as.numeric(format(date_time,"%m")),
    TimeShotHour = hour + minute/60,
    timeOfYear = (month - 1) * 1/12 + (day - 1)/365,   
    
    # add some dummy vars and create some other vars
    Country = "USA",
    Gear = "dummy",
    Quarter = "2")  %>%
  dplyr::mutate(across((c("Year", "Ship", "COMMON_NAME")), as.factor)) %>% 
  dplyr::select(wCPUE, GEAR_TEMPERATURE, BOTTOM_DEPTH, COMMON_NAME, EFFORT, 
                Year, Ship, Lon, Lat, lat, lon, sx, sy, 
                ctime, TimeShotHour, timeOfYear, Gear, Quarter, HaulDur, hauljoin)
```